Toward single particle reconstruction without particle picking: Breaking the detection limit

### 2018-08-27 pnas/Towards.tex
experimental details

In the first experiment, we estimated Einstein's image of size $50\times 50$ and mean zero from a growing number of micrographs, each of size $4096\times 4096$ pixels. A micrograph contains, on average, 700 occurrences of the target image at random locations. 
Thus, about 10\% of each micrograph contains signal. The micrographs are contaminated with additive white Gaussian noise with standard deviation $\sigma=3$ (this corresponds to SNR$\approx ?$). This high noise level is illustrated in Figure~\ref{fig:micro_example}. 
To simplify the experiment, we assume the number of signal occurrences and the noise standard deviation are known. Micrographs are generated such that any two occurrences are always separated by at least 49 pixels in each direction in accordance with~\eqref{eq:spacing}. %The separation restriction is discussed in more details in the Methods Section.

@@

\begin{figure}[h]
	\centering
	\includegraphics[scale=.68]{Einstein_progress} (left) power spectrum estimation error (right) RRR error after fixed number of 1000 iterations.
	%\caption{\label{fig:error_per_micro} Relative root mean square error of the estimate of Einstein's image as a function of the number of observed micrographs (logarithmic scale along both axes.)}
\end{figure}

@@

\TODO{Need to refer to this paper https://arxiv.org/abs/1806.00338}

>>>>

In the first experiment, we estimated Einstein's image of size $50\times 50$ and mean zero from a growing number of micrographs, each of size $4096\times 4096$ pixels. A micrograph contains, on average, 700 occurrences of the target image at random locations. 
Thus, about 10\% of each micrograph contains signal. The micrographs are contaminated with additive white Gaussian noise with standard deviation $\sigma=3$ (this corresponds to SNR$\approx1/375$). This high noise level is illustrated in Figure~\ref{fig:micro_example}. 
To simplify the experiment, we assume the number of signal occurrences and the noise standard deviation are known. Micrographs are generated such that any two occurrences are always separated by at least 49 pixels in each direction in accordance with~\eqref{eq:spacing}. %The separation restriction is discussed in more details in the Methods Section.

@@

\begin{figure}[h]
	\centering
	\includegraphics[scale=.68]{Einstein_progress} (left) power spectrum estimation error (right) RRR error after fixed number of 1000 iterations.
	\caption{\label{fig:error_per_micro} Relative root mean square error of the estimate of Einstein's image as a function of the number of observed micrographs (logarithmic scale along both axes.)}
\end{figure}

@@

\paragraph{2-D experiment.}

For the 2-D experiment shown in Figures~\ref{fig:Einst_example} and~\ref{fig:error_per_micro}, we generate $P$ micrographs of size $4096\times 4096$ pixels. 
In each micrograph, we place Einstein's image (of zero mean) of size $50\times 50$  in random locations, while preserving the separation condition~\eqref{eq:spacing}.  
This is done by randomly selecting $4000$ placements in the micrograph, one at a time with
an accept/reject rule based on the separation constraint and locations picked so far.
On average, $700$ images are placed in each micrograph.   
Then, i.i.d.\ Gaussian noise with standard deviation $\sigma=3$ is added, inducing an SNR of approximately $1/375$.
An example of a micrograph's excerpt is presented in the right panel of Figure~\ref{fig:micro_example}.
%Different micrographs are handled sequentially on a GPU, as GPUs are particularly well suited to execute simple instructions across large vectors of data. If multiple GPUs are available, segments can of course be handled in parallel.

In this experiment, we assume we know the noise level $\sigma$ and the total number of occurrences of the target image across all micrographs.
In stark contrast with the 1-D setup, the second-order autocorrelation determines almost any target image uniquely, up to reflection through the origin. This is because the second-order autocorrelations correspond to the Fourier magnitudes of the signal through the 2-D Fourier transform. 
Therefore, we estimate the signal's Fourier magnitudes (or power spectrum) from the Fourier magnitudes of the micrographs, at the cost of one 2-D fast Fourier transform (FFT) per micrograph. These can be computed highly efficiently and in parallel.

To recover the target image from the estimated power spectrum, we use a standard phase retrieval algorithm called relaxed-reflect-reflect (RRR). This algorithm iterates the map
\begin{align*}
z & \leftarrow z + \beta (P_2(2P_1(z) - z) - P_1(z))
\end{align*}
on an image $z$ of size $2L\times 2L$.
We set the parameter $\beta$ to 1.
The map is designed so that, if the estimated power spectrum is exact, then fixed points contain Einstein's image in the upper-left corner of size $L \times L$, possibly reflected through its origin, and zeros elsewhere. The operator $P_2(z)$ combines the Fourier phases of  the current estimation $z$ with the estimated Fourier magnitudes. The operator $P_1(z)$ zeros out all entries of $z$ outside the $L\times L$ upper-left corner. In all experiments, the algorithm halted after a fixed number of 2000 iterations.
%
%In order to compare the performance in multiple cases and at different noise levels,  and the iterate with the smallest error compared to the ground truth (up to the reflection ambiguity) is chosen as the solution. While this cannot be done in practice (since we do not have access to the ground truth to determine which iterate is best), this procedure enables us to compare a large number of instances in different noise environments. \TODO{Note the last two sentences!}

\paragraph{1-D experiment.}
For the 1-D experiment depicted in Figure~\ref{fig:1D_example}, we used a signal of length $L=21$ and generated an observation $y$ of length $N = 10(2L-1)M$, where $M$ is the number of repetitions. 
The observation was generated by randomly selecting placements in $y$, one at a time with an accept/reject rule based on the separation constraint and locations picked so far until reaching the desried number of signal occurences. Then, i.i.d.\ Gaussian noise with mean zero and standard deviation $\sigma = 3$ is added, to form the observed $y$. The resulting SNR of $y$ is about $1/175$.
%	 fix $K = 3$ signals of length $L = 21$. Following the forward model described at the beginning of this section, we generate an observation $y$ of length $12.3 \cdot 10^9$. Each of the three signals appears, respectively (and approximately), $30.0 \cdot 10^6$, $20.0 \cdot 10^6$ and $10.0 \cdot 10^6$ times in $y$ for a total of exactly $60 \cdot 10^6$ occurrences, such that at least $L-1$ zeros separate any two occurrences of any signals. 
%This is done by randomly selecting $60 \cdot 10^6$ placements in $y$, one at a time with an accept/reject rule based on the separation constraint and locations picked so far. For each placement, one of the three signals is picked at random according to the proportions $1/2, 1/3, 1/6$. Then, i.i.d.\ Gaussian noise with mean zero and standard deviation $\sigma = 3$ is added, to form the observed $y$. The resulting SNR of $y$
%% sqrt((m_want*sum(X.^2)')/(sigma^2*n))
%is about 1/9.

%This is enough noise to make cross-correlations of $y$ even with the true signals display peaks at essentially random locations, uninformative of the actual locations of the signal occurrences. Thus, we contend that it would be difficult for any algorithm to locate the signal occurrences, let alone to classify them according to which signal appears where.

%\TODO{TB: I would place the equation counting argument in the theory section in the paragraph of open questions (I marked th place)}
%
Given the observation $y$, we proceed to compute the autocorrelations. %The first-order autocorrelation is straightforward. 
We excluded biased elements of the autocorrelations; that is $a_y^2[\ell_1]$ and $a_y^3[\ell_1, \ell_2]$ such that $\ell_1, \ell_2$ or $\ell_1 - \ell_2$ are zero. 
This has the  practical effect that we need not know $\sigma$. 
%For second-order autocorrelations, notice from equation~\eqref{eq:data_ac} that $a_y^2[\ell]$ suffers no noise-induced bias for $\ell$ in $1$ to $L-1$. Thus, we omit $\ell = 0$, which has the practical effect that we need not know $\sigma$ to make sense of the computed quantities. Likewise, for third-order autocorrelations, $a_y^3[\ell_1, \ell_2]$ for $0 \leq \ell_1, \ell_2 \leq L-1$ such that $\ell_2 \leq \ell_1$ includes all relevant entries for our purpose (this accounts for symmetries), and we further exclude any such that $\ell_1, \ell_2$ or $\ell_1 - \ell_2$ are zero to avoid the need to estimate $\sigma$---
By also taking symmetries into account, the third-order autocorrelation contains $\frac{(L-1)(L-2)}{2}$ remaining entries. Thus, in total we have
\begin{align*}
1 + (L-1) + \frac{(L-1)(L-2)}{2} = \frac{1}{2} L (L-1) + 1
\end{align*}
coefficients. 
This redundancy suggests that it might be possible to estimate several signals simultaneously (compare with~\cite{boumal2017heterogeneous,bandeira2017estimation}).
%Since we aim to estimate $KL$ parameters (for the $K$ signals of length $L$) plus $K$ parameters (for the densities $\gamma_k$), an absolute upper bound on $K$ (simply to ensure we have at least as many equations as we have unknowns) is
%\begin{align*}
%K(L+1) \leq \frac{1}{2} L (L-1) + 1.
%\end{align*}
%Thus, $(L-1)/2$ (up to a small approximation) is an absolute upper limit on $K$ (compare with~\cite{boumal2017heterogeneous,bandeira2017estimation}). 
In practice, the autocorrelations are computed on disjoint segments of $y$ of length $100\cdot10^6$ (if the length of the measurement is larger than that) and added up, without correction for the junction points. Segments are handled sequentially on a GPU, as GPUs are particularly well suited to execute simple instructions across large vectors of data. If multiple GPUs are available, segments can of course be handled in parallel.

%Having computed the moments of interest, we now estimate signals $x_1, \ldots, x_K$ and coefficients $\gamma_1, \ldots, \gamma_K$ which agree with the data. We choose to do so by running an optimization algorithm on the following nonlinear least-squares problem:
Having computed the moments of interest, we now estimate the signal $x$ and coefficient $\gamma$ which agree with the data. We choose to do so by running an optimization algorithm on the following nonlinear LS problem:
%\begin{align}
%\min_{\substack{\hat x_1, \ldots, \hat x_K \in \mathbb{R}^{W} \\ \hat \gamma_1, \ldots, \hat \gamma_K > 0}}& w_1 \left( a_y^1 - \sum_{k=1}^K \hat \gamma_k a_{\hat x_k}^1 \right)^2 \\ & + w_2 \sum_{\ell = 1}^{L-1} \left( a_y^2[\ell] - \sum_{k=1}^K \hat \gamma_k a_{\hat x_k}^2[\ell] \right)^2  \nonumber \\&+ w_3 \sum_{\substack{2\leq\ell_1\leq L-1 \\ 1 \leq \ell_2 \leq \ell_1-1}} \left( a_y^3[\ell_1, \ell_2] - \sum_{k=1}^K \hat \gamma_k a_{\hat x_k}^3[\ell_1,\ell_2] \right)^2, \nonumber
%\label{eq:optim1D}
%\end{align}
\begin{align}
\min_{\substack{\hat x\in \mathbb{R}^{W} \\ \hat \gamma > 0}}& w_1 \left( a_y^1 - \hat \gamma a_{\hat x}^1 \right)^2  + w_2 \sum_{\ell = 1}^{L-1} \left( a_y^2[\ell] - \hat \gamma a_{\hat x}^2[\ell] \right)^2  \nonumber \\&+ w_3 \sum_{\substack{2\leq\ell_1\leq L-1 \\ 1 \leq \ell_2 \leq \ell_1-1}} \left( a_y^3[\ell_1, \ell_2] -  \hat \gamma a_{\hat x}^3[\ell_1,\ell_2] \right)^2, 
\label{eq:optim1D}
\end{align}
where $W \geq L$ is the length of the sought signals and the weights are set to $w_1 = 1/2, w_2 = 1/2n_2, w_3 = 1/2n_3$, where $n_2, n_3$ are the number of moments used: $n_2 = L-1$, $n_3 = \frac{(L-1)(L-2)}{2}$ (weights could also be set in accordance with variance estimates as in~\cite{boumal2017heterogeneous}).

Setting $W = L$ (as is a priori desired) is problematic because the above optimization problems appears to have numerous poor local optimizers. Similar phenomenon was recently observed in a related problem~\cite{zhang2018structured}.
Thus, we first run the optimization with $W = 2L-1$. This problem appears to have few poor local optima, perhaps because the additional degrees of freedom allow for more escape directions. Since we hope the signals estimated this way correspond to the true signals zero-padded to length $W$, we extract from each one a subsignal of length $L$ that has largest $\ell_2$-norm. This estimator is then used as initial iterate for~\eqref{eq:optim1D}, this time with $W = L$. We find that this procedure is reliable for a wide range of experimental parameters. To solve~\eqref{eq:optim1D}, we run the trust-region method implemented in Manopt~\cite{manopt}, which allows to treat the positivity constraints on coefficients $\hat \gamma$. Notice that the cost function is a polynomial in the variables, so that it is straightforward to compute it and its derivatives.

\paragraph{3-D experiments.}
